A Java Wrapper For Embarrassingly Parallel Problems

### 2013-11-25 paper/paper.tex
Beginning of Dan's paper revisions

A computation graph consists of nodes that represent the functions or processes. 
The edges of the graph show data being passed from one process to another. If 
the graph is disconnected, none of the processes need to communicate to complete
the computations. Such a computation is called "embarrassingly parallel." In a 
practical situation, there will be some inter-node communication to set up the 
problem, and then again to accumulate the results. Embarrassingly parallel 
problems can be easily performed on a cluster of servers or in any kind of 
distributed network, due to its lack of dependencies. The term embarrassingly 
parallel was coined by Cleve Moler in his 1986 paper "Matrix Computation on 
Distributed Memory Multiprocessors." \cite{history}

Programs in the field of Biology and Genetics often analyze long strings of
characters, whether they are DNA, RNA or protein sequences. In addition to
handling large file sizes, there is a lot of processing that goes into comparing
and manipulating strings. This time adds up when there are thousands of strings 
to process. However, these sequences often do not relate to each other and can 
be processed independently. Using an embarrassingly parallel approach, the 
processing time for all of the sequences can often be cut down exponentially 
with very minimal effort by the programmer.

The framework provides a "wrapper" for the original program that mimics the
behavior of the original program. It calls the serial programs on multiple Java 
threads. Each thread acts like the serial application, reading a subset of the 
input the same and creating the output in the same format. This paper describes
a framework that makes it easy to parallelize an potentially embarrassingly
parallel program that reads from a file, writes to a file and whose input can be
broken up into work units by a regular expression. The framework gives an
exponential decrease in runtime as more processor help in the computation. This
decrease levels off between 50 and 100\% CPU usage due to the other tasks on the
machine that take up memory.

@@

The motivation for this effort was research into different forms of the 
species \emph{Cicer arietinum}, the chickpea. The two main forms, Kabuli and 
Desi, are very different in size and color. The Desi is also more resistant to 
diseases and are more effective at root nodule symbiosis. The Kabuli has a large 
but soft shell, which make it at better food source for humans. These two types 
are both cultivated and have been under human selection for several thousand 
years. We wanted to use a Ka/Ks equation to quantify the evolution and selection 
of these chickpeas. We found a Ka/Ks program written by Zhang et al. \cite{kaks}
but had three variants that we wanted to analyze, each with thousands of 
sequences. The Ka/Ks Calculator program takes somewhere from five to thirty 
seconds to process one pair of sequences.

Instead of attempting to rewrite the preexisting program, we decided it would be
better to look at other options to process the sequences faster. Trying to 
parallelize the serial Ka/Ks program could present several problems: introducing
bugs, breaking the algorithm, dealing with the learning curve of C++ threading. 
We decided that a  different approach was necessary, mostly due to the short 
time given to produce results. A simple Java program was written to break up the
input and spawn threads that would run the Ka/Ks Calculator process. When the 
input was calculated, the contents of the output files were concatenated 
together to produce a final output that matches the format and order that would 
have been produced by the Ka/Ks Calculator program.

%TODO: get a better source? Talk about UCLUST and octupus
The Ka/Ks Calculator is not the only embarrassingly parallel program that has the
potential to take several days to run on a data set. In fact, there are several
embarrassingly parallel programs in the biological sciences. To deal with large
amounts of data, the best method is often to group them into clusters. 
Hierarchical clustering algorithms take the data and then organize into
hierarchical structure, usually a dendrogram, based on a priority matrix. There 
are hierarchical clustering algorithms that can be implemented in parallel using
an embarrassingly parallel approach.  With several applications that would 
benefit from parallelism in this field, it was obvious it would be desirable to 
develop a framework for creating wrapper programs. \cite{cluster}

@@

The other purpose of the ChunkManager class is to combine the resulting Chunks 
back together when the worker threads have finished. Once the work has been
done, all of the separate results are spread across several files. Those results 
must be retrieved and combined to construct the final output file. This output 
file must look the same as the output file would look if the program were run 
serially; the output must come in the same order as the inputs in the input file
and headers may need to be deleted from the separate files and added to the 
combined file. There are two other options if the end-user does not want this 
behavior. They can choose to write their own merger in Java, using the 
CustomMerge class, or write their own script to do the merging.

As more features are added to the framework and as more programs start to use 
the framework, there are more choices for the end-user to make with regards to 
how they want their program to execute. The number of threads to use, various 
flags for the executable and the statistics output are just a few choice the 
user can make. A start-up wizard is displayed at runtime to systematically allow
the end-user to choose the settings for the run. The ConfigWizard class brings 
the user through this process and creates an instance of Configuration when it 
is completed to save the user's settings.

Lastly, the worker threads will keep records during execution. The threads 
records their uptime and the number of pieces of work they execute. The runtime 
for each piece of work is also recorded. This information will be used by us to
evaluate the effectiveness of our approach, but can also provide useful
information the can help a user tine a particular application on a particular
type of input data. The statistics for each thread can be found in the Worker 
subclass and the individual Chunk objects will store their own statistics about 
runtime and size.

@@

In order to observe how the framework changes the runtime of sequential programs 
when using varying numbers of threads, we designed a simple experiment. The 
Ka/Ks Calculator was run through the framework with ten different samples, each 
containing sixteen units of work. Each sample was run by the framework at five 
different level of numbers of worker threads, starting at 1 and increasing by 
powers of two until the 100\% CPU usage (16 processors). These tests were run
with minimal other activity on the server.

@@

@@

As the numbers of threads increased, the runtime of the program decreased 
exponentially. In a few cases, there was an increase in time when attempting to 
max out the CPU usage. This occurred because the framework had to wait for the 
CPU to be finished with its previous work before it was allowed to start any 
further jobs. In the cases when the framework executed faster while the CPU was 
maxed out, it was only faster a minute at the most. However, running the 
framework at 50\% of the CPU is more efficient than running using 100\% of the 
CPU. The runtime drop achieved when using 100\% of the CPU versus 50\% is not 
large enough to warrant using all of the machine's power.

Using Java was a good choice for implementation ease and future implementation
extension by others, but it may not produce the fastest results. Languages
like Hadoop, Clojure or other multithread oriented language may have been a
better choice due to their parallel nature. In addition to a single machine 
working to process the data quickly, we can also look to the cloud for help by 
using a networked solution like MPI or a cloud solution like Amazon instances. 
The MapReduce framework works similarly to our framework, there is a step that
splits the data, the Map function is the user's executable and the merge step is
the Reduce function in the framework. In the future, we would like to see this 
work be split into a local front-end and a cloud back-end. This allows for an 
even greater runtime decrease because the number of compute nodes available is 
larger than the number of processors on most machines. Splitting the framework 
also allows for better testing of cloud environments, where the cloud 
environment itself could be a configurable option for end users.

>>>>

A computation graph consists of nodes that represent the functions or processes.
The edges of the graph show data being passed from one process to another. If
the graph is fully disconnected, none of the processes need to communicate to
complete the computations. Such a computation is called embarrassingly
parallel.\footnote{The term embarrassingly parallel was coined by Cleve Moler in
his 1986 paper ``Matrix Computation on Distributed Memory Multiprocessors.''} In
a practical situation, there will be some inter-node communication to set up the
problem, and then again to accumulate the results.  Embarrassingly parallel
problems can be easily performed on a cluster of servers or in any kind of
distributed network, due to its lack of dependencies.

Programs in the field of Biology and Genetics often analyze long strings of
characters, whether they are DNA, RNA or protein sequences. In addition to
handling large file sizes, there is a lot of processing that goes into comparing
and manipulating strings. This time adds up when there are hundreds of thousands
of strings to process. However, these sequences often do not relate to each
other and can be processed independently. Using an embarrassingly parallel
approach, the processing time for all of the sequences can often be cut down
exponentially with very minimal effort by the programmer.

This paper describes a framework that makes it easy to parallelize a potentially
embarrassingly parallel program that reads from a file, writes to a file and
whose input can be broken up into work units in an easily identifiable manner.
The framework provides a ``wrapper'' for the original program that mimics the
behavior of the original program. It calls the serial programs on multiple Java
threads, such that each thread acts like the serial application, reading a
subset of the input in the same manner and creating the output in the same
format. The framework can result in a significant exponential decrease with very
little programmer effort.

@@

TODO: Arrows?
The motivation for this effort was based on research into different forms of the
species \emph{Cicer arietinum}, the chickpea. The two main forms, Kabuli and
Desi, are very different in size and color. The Desi is also more resistant to
diseases and are more effective at root nodule symbiosis. The Kabuli has a large
but soft shell, which make it at better food source for humans. Both forms are
cultivated and have been under human selection for several thousand years. We
wanted to use an existing Ka/Ks program (Zhang et al. \cite{kaks}) to quantify
the evolution and selection of these chickpeas. We needed to analyze three
variants, each with thousands of sequences. The Ka/Ks Calculator program took
between five to thirty seconds to process one pair of sequences on a commodity
desktop.

At this rate, we predicted it would take ten days of constant computation to
complete the processing we needed. We neither had the time nor the expertise to
develop a new parallel Ka/Ks program. We were also reluctant to revise the
existing code to to implement the parallel computation because of the danger of
introducing bugs into unfamiliar code and because of the learning curve required
for C++ threading. Instead we wrote, a simple Java program to break up the input
and spawn threads that would run the Ka/Ks Calculator process without having to
make any changes to the code. When al of the threads ended, the contents of the
output files were concatenated together to produce a final output that matches
the format and order that would have been produced by a single invocation of the
Ka/Ks Calculator program.

%TODO: get a better source? Talk about UCLUST and octupus
Ka/Ks Calculator is not the only embarrassingly parallel program that has the
potential to take several days to run on a data set. In fact, there are several
embarrassingly parallel programs in the biological sciences. To deal with large
amounts of data, the best method is often to group them into clusters.
Hierarchical clustering algorithms take the data and then organize into
hierarchical structure, usually a dendrogram, based on a priority matrix. There
are hierarchical clustering algorithms that can be implemented in parallel using
an embarrassingly parallel approach.  With several applications that would
benefit from parallelism in this field, it was obvious it would be desirable to
develop a framework for the  parallelization task. \cite{cluster}

@@

The other purpose of the ChunkManager class is to combine the resulting Chunks
back together when the worker threads have finished. Once the work has been
done, all of the separate results are spread across several files. Those results
must be retrieved and combined to construct the final output file. This output
file must look the same as the output file would look if the program were run
serially; the output may need to come in the same order as the inputs in the
input file and headers may need to be deleted from the separate files and added
to the combined file. End-users can also choose to write their own merge code in
Java, using the CustomMerge class, or write their own independent merge script.

The end-user can control many aspects of the parallelization of the code,
including the number of threads to use, various flags for the executable and the
statistics output.  A start-up wizard allows the end-user to choose the settings
for the run and creates an instance of Configuration when it is completed to
save the user's settings, allowing the user to rerun with the same settings
without having to go through the wizard again.

Lastly, the worker threads keep records during execution. The threads record
their total uptime and the number of Chunks they execute, and the runtime for
each piece of work. This information will be used to evaluate the effectiveness
of the approach, but can also provide useful information the can help a user
tine a particular application on a particular type of input data. The statistics
for each thread can be found in the Worker subclass and the individual Chunk
objects will store their own statistics about runtime and size.

@@

We evaluated the convenience and performace of the framework on two different
applications: the Ka/Ks Calculator and two different sequence clustering
algorithms. In all cases, we compared the framework based application with the
original using different data partitions and different thread counts.

\subsection{Ka/Ks Calculator}
Ka/Ks Calculator was run through the framework with ten different samples, each 
containing sixteen units of work. Each sample was run by the framework thread
counts, starting at 1 and increasing by 
powers of two until the 100\% processor usage (16 processors). These tests were run
with minimal other activity on the server.

@@

Figure 2 summarized the results; each line shows the time taken for each chunk
of the data.
As the numbers of threads increases, the runtime of the program decreases 
exponentially. In a few cases, there was an increase in time when attempting to 
max out the CPU usage. This occurs because the framework has to wait for the 
CPU to be finished with its previous work before it can start any 
further jobs.

TODO(Keep?): In the cases when the framework executed faster while the CPU was 
maxed out, it was only faster a minute at the most. However, running the 
framework at 50\% of the CPU is more efficient than running using 100\% of the 
CPU. The runtime drop achieved when using 100\% of the CPU versus 50\% is not 
large enough to warrant using all of the machine's power.

\subsection{uClust}

Place holder text

\subsection{Octupus}

Place holder text

@@

TODO: First paragrpah should summarize performce: its clear that framework
provices simple and effective module for adding parallelism to a class of
file-oriented EP apps. More self praise.

Using Java was a good choice for implementation ease and future implementation
extension by others, but it may not produce the fastest results. Languages like
Hadoop, Clojure or other multithread oriented language may have been a better
choice due to their inherit parallel nature. In addition to a single machine
working to process the data quickly, we can also look at distributed solutions
by using MPI or a cloud solution like Amazon instances. The MapReduce framework
works similarly to our framework, there is a step that splits the data, the Map
function is the user's executable and the merge step is the Reduce function in
the framework.

In the future, we would like to see this work be split into a
local front-end and a cloud back-end. This allows for an even greater runtime
decrease because the number of compute nodes available is larger than the number
of processors on most machines. Splitting the framework also allows for better
testing of cloud environments, where the cloud environment itself could be a
configurable option for end users.
TODO: MORE
